Variant calling with transformers

### 2022-05-28 manuscript/manuscript.tex
Manuscript updates

@@

The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position. The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.

@@

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions. We additionally included `true negative' regions where no known variant was present. The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.

For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.

@@

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.  While we obtain satisfactory results with this approach, two issues were apparent. First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes. Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes. Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.

@@

Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region. Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant. Positions closer than 100bp are merged into a single region.

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.

\subsection{Variant quality calculation and false positive removal}

The above variant detection procedure has two shortcomings. First, if every detected variant is emitted, then specificity is poor. Second, variant quality scores are poorly calibrated and do not closely reflect the probability that a variant is real. To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).  The resulting score is used as a final variant quality score that can be used for tuning the tradeoff between sensitivity and specificity.

@@

\subsection{Complex regions}
 While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly of reads to identify the true haplotype. We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings in addition to extensive read soft-clipping. 
 To identify such regions we first obtain all indel variants from Genome-in-a-Bottle greater than X bp. We then examine the reads in the alignment file and flag any region that contains more than Y distinct indel start sites or at least Z reads with more than 10bp softclipped. We then

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping. In addition, the model learns to employ local context and read-specific features to predict phase

 Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.

For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work the exploration of the many transformer variations proposed in recent years (vision transformers, swin transformers, fast-former, etc.)

@@

@@

@@

@@

>>>>

While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al. (2021) for generation of accurate consensus reads from PacBio data. The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region. In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.

@@

The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs. The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.

@@

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions. We additionally included `true negative' regions where no known variant was present.  For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.  We examine three different sizes of training data sets, with 5.5M, 10.4M, and 20.4M examples.

\subsection{Training procedure}

Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$.

@@

We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.  While we obtain satisfactory results with this approach, two issues were apparent. First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes. Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes. Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.

@@

Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist. Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant. Positions closer than 100bp are merged into a single region.

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the distance of the variant from the window start coordinate.

\subsection{Variant quality calculation}

The above variant detection procedure has two shortcomings. First, if every potential variant is emitted, then specificity is poor. Second, the model does not innately produce well calibrated variant quality scores. To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).  The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.

The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb. True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al. 2015). We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and 'balanced' class weighting option.

@@

\subsection{Complex regions}
 While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly to identify the true haplotype. We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings or extensive read soft-clipping.

 To identify variants requiring complex reconstruction, we first obtain all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22. For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp softclipped as requiring 'complex' reconstruction. These variants are moved to a separate truth set. After running our variant calling procedure on the alignment file, we compute the fraction of variants from the truth set that were identified by our model. This procedure was repeated for all 17 samples in our data set.

 Our model is able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).  Although model performance does not match current state-of-the-art callers, it nonetheless is able to reconstruct the true haplotype in most cases. Overall, these results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads,

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads. On short read WGS data, our approach demonstrates performance similar to current top-performing callers, with somewhat higher sensitivity but lower precision for insertion/deletion variants ()

Our analysis demonstrates that the model learns many capabilities of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping. While our model's sensitivity is ~5\% less than HaplotypeCaller for variants in these complex regions, we nonetheless are able to reconstruct the correct haplotypes in more than 90\% of cases.   In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.

All models in this work were trained using a standard cross-entropy loss function.  Cross-entropy loss is probably not optimal for variant detection, because insertion and deletion errors are more severely punished than SNVs, and the loss has a strong dependence on the relative position of the variant in the input sequence.  During development we experimented with the differentiable Smith-Waterman loss function as described by Petti et al. (2021), but this lead to reduced sensitivity in our experiments.

For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g. Dosovitskiy 2018, Fedus et al. 2021, Liu et al 2021, Wu et al. 2021). Our simple decoder only predicts haplotypes that are equal in length to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.  In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.

@@

\vspace{8pt}
Baid, Gunjan, et al. "DeepConsensus: Gap-Aware Sequence Transformers for Sequence Correction." bioRxiv (2021).

\vspace{8pt}
Cleary, John G., et al. "Comparing variant call files for performance benchmarking of next-generation sequencing variant calling pipelines." BioRxiv (2015): 023754.

@@

\vspace{8pt}
Dosovitskiy, Alexey, et al. "An image is worth 16x16 words: Transformers for image recognition at scale." arXiv preprint arXiv:2010.11929 (2020).

\vspace{8pt}
Fedus, William, Barret Zoph, and Noam Shazeer. "Switch transformers: Scaling to trillion parameter models with simple and efficient sparsity." arXiv preprint arXiv:2101.03961 (2021).

@@

\vspace{8pt}
Petti, Samantha, et al. "End-to-end learning of multiple sequence alignments with differentiable Smith-Waterman." BioRxiv (2021).

@@

\vspace{8pt}
Wu, Chuhan, et al. "Fastformer: Additive attention can be all you need." arXiv preprint arXiv:2108.09084 (2021).
